---
title: "Description of data"
format: html
---

# 1. Load libraries

```{r}
#| message: false
#| warning: false
library("tidyverse")
```

# 2. Load data

Loading the data saved in the 03_dat_aug.tsv file

```{r}
genome_metadata_aug <- readr::read_tsv("../data/03_dat_aug.tsv")
```

# 3. Data overview

## 3.1 Structure of the dataset

Using the str command to investigate, what the data looks like:

```{r}
glimpse(genome_metadata_aug)
```

From the overview of the data structure, there are 618 rows and 21 columns. The data Includes assembly metrics, rRNA/tRNA counts, taxonomy, and geography.

## 3.2 Missing values in key variables

We summarize missing values for the data fields to see where there are most missing values.

```{r}
genome_metadata_aug |>
  summarise(across(everything(), ~ sum(is.na(.x)))) |>
  pivot_longer(
    everything(),
    names_to = "variable",
    values_to = "n_missing"
  ) |>
  arrange(desc(n_missing)) |>
  slice_head(n = 6)
```

There are four variables with missing data in the cleaned dataset. *Species* has by far the most missing values (138), followed by *Country* and *Continent* (56 each), while *Genus* has only a small number of missing entries (6). All other variables are fully complete.

# 4. Taxonomic composition

## 4.1 Phylum counts

Counting the different Phylums in the dataset

```{r}
genome_metadata_aug |>
count(Phylum, sort = TRUE)
```

## 4.2 Barplot of phyla

```{r}
genome_metadata_aug |>
count(Phylum) |>
ggplot(aes(x = fct_reorder(Phylum, n), y = n)) +
geom_col() +
coord_flip() +
labs(
x = "Phylum",
y = "Number of genomes",
title = "Number of genomes per phylum"
)
```

From the plit it is seen that the data contains MAGs from several bacterial phyla commonly found in the vaginal microbiome.\
The most abundant phyla include Bacillota, Actinobacteriota, and Bacteroidota, while others such as Synergistota appear less frequently.\

# 5. Genome quality

## 5.1 Completeness and contamination

Barplot af inddeling af forskellige Phylum

Hvordan er datasammensat

Hordan er data sammen sat geographic (abundance)

Hvilke typer af bakterier ser man mest på continent/land

## 5.2 Quality categories (not created yet)
