---
title: "08_analysis_4"
format: html
---

# Loading libraries

```{r}
library(tidyverse)
library(pheatmap)
library(broom)
library(dplyr)
library(ggplot2)
library(ggrepel)
```

# Preparing dataset

## Removing NA values for "Countries"

From the augmented data, we'll remove the genomes without country information.

```{r}
genomes_country <- genome_metadata_aug %>%
  filter(!is.na(Country))
```

# Functions

## Function for preparing data and normalizing it based on the dataset size

```{r}
normalize <- function(data, tree_group){
  
  group_freq <- genomes_country %>%
    group_by(Country, .data[[tree_group]]) %>%
    summarise(count = n(), .groups = "drop") %>%
    left_join(genomes_country %>%
                count(Country, name = "n_samples")) %>%
    mutate(freq_norm = count / n_samples) %>%
    select(Country, .data[[tree_group]], freq_norm) %>%
    pivot_wider(names_from = .data[[tree_group]],
                values_from = freq_norm,
                values_fill = 0)
}
```

# Heatmap

Defining a function to produce the heatmap.

```{r}

# Defining function for heatmap based on species, genus, family etc...
heatmap_abundance_per_country <- function(tree_group){
  
  # Check that the column exists
  if (!tree_group %in% names(genomes_country)) {
    stop(paste("Column", tree_group, "not found in genomes_country"))
  }
  
  group_freq <- normalize(data = genomes_country, tree_group = tree_group)
  
  # Convert to matrix, using Country as rownames
  group_freq_mat <- group_freq %>%
    column_to_rownames("Country") %>%
    as.matrix()
  
  # Remove zero-variance columns
  group_freq_mat <- group_freq_mat[, apply(group_freq_mat, 2, sd) > 0, drop = FALSE]
  
  # alternative without scaling:
  pheatmap(group_freq_mat,
           cluster_cols = TRUE,
           scale = "none",
           main = paste0(tree_group, " frequency per country (normalized)")
  )
}

```

Using the function to produce heatmap

```{r}
heatmap_family <- heatmap_abundance_per_country(tree_group = "Family")
heatmap_order <- heatmap_abundance_per_country(tree_group = "Order")
heatmap_phylum <- heatmap_abundance_per_country(tree_group = "Phylum")
```

## Saving results

```{r}

ggsave(
  filename = "../results/Heatmap_family.png",
  plot = heatmap_family,
  width = 190,
  height = 100,
  units = "mm",
  dpi = 300
)
ggsave(
  filename = "../results/Heatmap_order.png",
  plot = heatmap_order,
  width = 190,
  height = 100,
  units = "mm",
  dpi = 300
)
ggsave(
  filename = "../results/Heatmap_phylum.png",
  plot = heatmap_phylum,
  width = 190,
  height = 100,
  units = "mm",
  dpi = 300
)
```

# Principal Component Analysis

```{r}
PCA_abundance_per_country <- function(tree_group) {
  

  group_freq <- normalize(data = genomes_country, tree_group = tree_group)
  
  # Extract country names
  country_names <- group_freq$Country
  
  # Remove Country column and run PCA on numeric data only
  pca_fit <- group_freq %>% 
    select(-Country) %>%  # Remove the Country column
    scale() %>%            # scale data
    prcomp()               # do PCA
  
  # Get variance explained
  variance_explained <- pca_fit %>%
    tidy(matrix = "eigenvalues")

  # Get rotation matrix (loadings)
  rotation_matrix <- pca_fit %>%
    tidy(matrix = "rotation")
  
  # Get PCA scores and add country names back
  pca_scores <- pca_fit %>%
    augment(group_freq %>% select(-Country)) %>%
    mutate(Country = country_names)
  
  # Calculate % variance for axis labels
  pc1_var <- round(variance_explained$percent[1] * 100, 1)
  pc2_var <- round(variance_explained$percent[2] * 100, 1)
  
  # Prepare loadings for biplot (scale them for visualization)
  loadings_biplot <- rotation_matrix %>%
    filter(PC %in% c(1, 2)) %>%
    pivot_wider(names_from = PC, values_from = value, names_prefix = "PC") %>%
    mutate(
      PC1 = PC1 * 3,  # Scale factor for visibility (adjust as needed)
      PC2 = PC2 * 3
    )
  
  # Plot 1: PCA biplot with countries AND species loadings
  p1 <- ggplot() +
    # Add country points and labels
    geom_point(data = pca_scores, 
               aes(x = .fittedPC1, y = .fittedPC2, color = Country), 
               size = 4,
               show.legend = TRUE) +
    geom_text(data = pca_scores, 
              aes(x = .fittedPC1, y = .fittedPC2, label = Country, color = Country),
              vjust = -1,
              size = 4,
              show.legend = FALSE) +
    # Add loading arrows
    geom_segment(data = loadings_biplot,
                 aes(x = 0, y = 0, xend = PC1, yend = PC2),
                 arrow = arrow(length = unit(0.3, "cm")),
                 color = "gray50",
                 linewidth = 0.7) +
    geom_text_repel(data = loadings_biplot,
                    aes(x = PC1, y = PC2, label = column),
                    color = "gray10",
                    size = 3.5,
                    fontface = "italic",
                    box.padding = 0.5,      # Space around labels
                    point.padding = 0.3,    # Space from arrow tips
                    segment.color = "gray60",  # Connection lines
                    segment.size = 0.3,
                    max.overlaps = Inf) +   # Allow all labels
    labs(title = paste0("PCA Biplot: Countries + ", tree_group),
         x = paste0("PC1 (", pc1_var, "%)"),
         y = paste0("PC2 (", pc2_var, "%)")) +
    theme_minimal() +
    theme(legend.position = "right",
          plot.title = element_text(face = "bold"))
  
  
  # Plot 2: Scree plot (variance explained)
  p2 <- ggplot(variance_explained, aes(x = PC, y = percent)) +
    geom_col(fill = "steelblue", alpha = 0.8) +
    scale_y_continuous(labels = scales::percent_format()) +
    labs(title = "Variance Explained by Each PC",
         x = "Principal Component",
         y = "Variance Explained (%)") +
    theme_minimal() +
    theme(plot.title = element_text(face = "bold"))
  

  # ggsave
  ggsave(filename = "../results/PCA_biplot.png",
       plot = p1,
       width = 100,
       height = 70,
       units = "mm")
    
  ggsave(filename = "../results/PCA_variance_explained.png",
         plot = p2,
         width = 100,
         height = 70,
         units = "mm")
  
}

```

Running the function

```{r}

# Run the function
PCA_abundance_per_country(tree_group = "Phylum")

```

## 

# Tests: Species' distribution between counties

Simply counting the species (or another tree_group) per country.

```{r}

counting_species <- function(data, tree_group){
  data |> 
  group_by(Country, .data[[tree_group]]) |>
  summarise(count = n(), .groups = "drop") |> 
  pivot_wider(names_from = .data[[tree_group]],
              values_from = count,
              values_fill = 0)
}

```

## Chi-squared test, with permutation.

Test if the distribution of species differs between countries. We use simulate.p.value = TRUE to compute the p-value via Monte Carlo simulation, which is more reliable when some expected counts are small or zero.

### For Species

```{r}
data <- counting_species(data = genomes_country, tree_group = "Species")

tbl <- as.matrix(data[ , -1])
rownames(tbl) <- data$Country

chisq.test(tbl, simulate.p.value = TRUE, B = 1e6)

```

The difference in distribution of species between countries is not significant at all, as the p-value is 1.

### For Family

```{r}
data <- counting_species(data = genomes_country, tree_group = "Family")

tbl <- as.matrix(data[ , -1])
rownames(tbl) <- data$Country

chisq.test(tbl, simulate.p.value = TRUE, B = 1e6)

```

The difference in distribution of species between countries is not significant at all, as the p-value is 0.8.

### For Phylum

```{r}
data <- counting_species(data = genomes_country, tree_group = "Phylum")

tbl <- as.matrix(data[ , -1])
rownames(tbl) <- data$Country

chisq.test(tbl, simulate.p.value = TRUE, B = 1e6)

```

The difference in distribution of species between countries is not significant, as the p-value is 0.6

# Endometriosis chi-square test

To assess whether

Check that the data contains the variable of interest:

```{r}
names(genomes_country)
```

Creating a table that contains the information about the country and the endometriosis association:

```{r}
tab <- table(
  Country = genomes_country$Country,
  Endometriosis = genomes_country$endometriosis_associated
)
tab
prop.table(tab, margin = 1)
```

Performing a chi-square test to see

```{r}
chi_res <- chisq.test(tab)

chi_res
chi_res$expected  # expected counts under H0

```

```{r}
fisher_res <- fisher.test(tab)

fisher_res

```

```{r}
pairwise.prop.test(
  x = table(genomes_country$Country, genomes_country$endometriosis_associated),
  p.adjust.method = "holm"
)


```

```{r}
countries <- c("USA", "China")

tab <- table(
  genomes_country$Country[genomes_country$Country %in% countries],
  genomes_country$endometriosis_associated[genomes_country$Country %in% countries]
)

tab
fisher.test(tab)

```

```{r}
countries <- c("USA", "Denmark")

tab <- table(
  genomes_country$Country[genomes_country$Country %in% countries],
  genomes_country$endometriosis_associated[genomes_country$Country %in% countries]
)

fisher.test(tab)
```

```{r}
countries <- c("Denmark", "China")

tab <- table(
  genomes_country$Country[genomes_country$Country %in% countries],
  genomes_country$endometriosis_associated[genomes_country$Country %in% countries]
)

fisher.test(tab)
```
