---
title: "Sigrid 2"
format: html
output-dir: results
---

# Analysis X

1)  Endobiota Study 2019, "The endobiota study: comparison of vaginal, cervical and gut microbiota between women with stage 3/4 endometriosis and healthy controls" (<https://www.nature.com/articles/s41598-019-39700-6>)

Across these vaginal datasets, the most consistently reported endometriosis-associated genera are:

-   Anaerococcus

-   Ureaplasma

-   Gardnerella

-   Veillonella

-   Corynebacterium

-   Peptoniphilus

-   Candida albicans

-   Alloscardovia

## Load libraries

```{r}
#| message: false
#| warning: false
library("tidyverse")
```

## Load data

```{r}
genome_metadata_aug <- readr::read_tsv("../data/03_dat_aug.tsv")
```

## 1. Annotating MAGs with endometriosis-associated genera.

We add a new column to the MAG dataset that flags whether each MAG belongs to a genus repeatedly associated with endometriosis in the vaginal microbiome. Labeling True or False

```{r}

endo_genera <- c(
  "Anaerococcus",
  "Ureaplasma",
  "Gardnerella",
  "Veillonella",
  "Corynebacterium",
  "Peptoniphilus",
  "Candida albicans", 
  "Alloscardovia"
)


genome_metadata_aug <- genome_metadata_aug |>
  dplyr::mutate(
    endometriosis_associated = Genus %in% endo_genera
  )


genome_metadata_aug |>
  dplyr::count(endometriosis_associated)

```

There are 542 MAGs marked a non-endometriosis associated and 76 MAGs marked as endometriosis associated.

## 2. GC content of endometriosis-associated genera

We compare GC content between associated and non-associated groups.

```{r}
genome_metadata_aug |>
  ggplot2::ggplot(aes(x = endometriosis_associated, y = GC_content)) +
  ggplot2::geom_boxplot() +
  ggplot2::labs(
    x = "Endometriosis-associated genus",
    y = "GC content (%)",
    title = "GC content of MAGs by endometriosis association"
  )

```

Overall, the two groups show overlapping GC-content distributions, indicating no major genomic differences in GC composition between endometriosis-associated and non-associated taxa. MAGs annotated as endometriosis-associated span a fairly wide GC range, and their median GC content is slightly lower than that of non-associated genera, but the difference is modest and not indicative of a distinct GC signature. This suggests that GC content is not a distinguishing genomic feature of vaginal taxa associated with endometriosis, and that their relevance is likely driven by ecological or abundance differences rather than genome composition.

## 3. Genome length comparison

We examine whether MAGs associated with endometriosis differ in genome size.

```{r}
genome_metadata_aug |>
  ggplot2::ggplot(aes(x = endometriosis_associated, y = Length)) +
  ggplot2::geom_boxplot() +
  ggplot2::scale_y_log10() +
  ggplot2::labs(
    x = "Endometriosis-associated genus",
    y = "Genome length (bp, log scale)",
    title = "Genome length of MAGs by endometriosis association"
  )

```

MAGs from endometriosis-associated genera tend to have slightly larger genome sizes compared with other vaginal taxa, although the ranges of both groups overlap broadly. This suggests that genome size is not strongly structured by endometriosis association, but the associated genera may include species with somewhat more expansive genomes.

## 4. Completeness and contamination

We assess whether endometriosis-associated genera differ in assembly quality.

### 4.1 Completeness

```{r}

genome_metadata_aug |>
  ggplot2::ggplot(aes(x = endometriosis_associated, y = Completeness)) +
  ggplot2::geom_boxplot() +
  ggplot2::labs(
    x = "Endometriosis-associated genus",
    y = "Completeness (%)",
    title = "MAG completeness vs endometriosis association"
  )


```

Completeness values are high for both groups, with only minor variation. Endometriosis-associated MAGs show completeness levels similar to those of other genera, suggesting that their assemblies are of comparable quality.

### 4.2 Contamination

```{r}

genome_metadata_aug |>
  ggplot2::ggplot(aes(x = endometriosis_associated, y = Contamination)) +
  ggplot2::geom_boxplot() +
  ggplot2::labs(
    x = "Endometriosis-associated genus",
    y = "Contamination (%)",
    title = "MAG contamination vs endometriosis association"
  )
```

Contamination levels are consistently low across both groups, with no clear differences between endometriosis-associated and non-associated MAGs. This indicates that assembly quality is comparable, and any genomic differences observed are unlikely to be driven by differences in contamination.

Using genera previously linked to endometriosis in vaginal microbiome studies, we annotated a subset of MAGs and compared their genomic features to the rest of the dataset. Across GC content, genome size, completeness, and contamination, the two groups showed only minor differences with broad overlap. This indicates that endometriosis-associated genera do not form a genomically distinct group within the vaginal microbiome. Instead, their relevance to endometriosis likely reflects differences in abundance or community composition rather than major differences in genome architecture.
